An examination of the accessibility implications of a pilot COVID-19 vaccination program in Hamilton, Ontario

### 2021-04-06 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.Rmd
v1 Final edits before submission to Transport Findings

Research Questions and Hypotheses
==========================

@@

Critics were swift to point out that the list of pharmacies approved for Hamilton by the province were mostly located in lower density parts of the city that are not well serviced by transit and are difficult to reach by foot\footnote{See inter alia: \url{https://twitter.com/RyanMcGreal/status/1378027149790224386?s=20} and \url{https://twitter.com/NrinderWard3/status/1378679195514060801?s=20}.}. Indeed, as seen in Figure \ref{fig:pharmacies-and-regions}, a vast majority of the pharmacies are in suburban Hamilton. The issue is somewhat less clear-cut when we consider that Hamilton's population skews suburban (see Figure \ref{fig:population-map}). Given the target demographic for the program, it is possible that suburban sites could be convenient for mature adults and young old: the population aged 55 to 69 in Hamilton is approximately `r prettyNum(round((table_comparison[7,2]/1000), 2) * 1000, big.mark = ",")` suburban, `r prettyNum(round((table_comparison[8,2]/1000), 2) * 1000, big.mark = ",")` urban, and only `r prettyNum(round((table_comparison[6,2]/1000), 2) * 1000, big.mark = ",")` rural. Nevertheless, the selection of sites by the province raises some important questions\footnote{The decision-making process to select these sites appears to have been opaque, and the Mayor of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}. As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may introduce significant inequities in access.

@@

```{r pharmacies-map, echo=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:pharmacies-and-regions}Regions with the City of Hamilton; the location of pharmacies in pilot is shown (black triangles) and urban locations for scenario analysis (white circles)"}
#map_pharmacies <-
ggplot() +
  geom_sf(data = urban_types,
          aes(fill = Type),
          color = "lightgray") +
  geom_sf(data = pharmacy_locations,
          aes(shape = Type,
              color = Type),
          size = 2.5,
          show.legend = FALSE) +
  scale_color_manual(values = c("Candidate" = "white",
                                "Pilot" = "black")) +
  scale_fill_manual(values = c("Urban" = "firebrick2",
                               "Suburban" = "cornflowerblue",
                               "Rural" = "forestgreen")) +
  guides(fill = guide_legend(title = "")) + 
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 7),
        legend.position = "bottom",
        legend.text = element_text(size = 8))
```

```{r population-map, echo=FALSE, warning=FALSE, fig.height=8, fig.align="center", fig.cap="\\label{fig:population-map}Distribution of population age 55+ in the City of Hamilton"}
ggplot() +
  geom_sf(data = data_da_2016 %>%
            filter(`Region Name` == "Hamilton") %>%
            dplyr::select(Population_55to69, Population_70plus) %>%
            pivot_longer(cols = -geometry, names_to = "Age", values_to = "Population") %>%
            mutate(Age = factor(Age,
                                levels = c("Population_55to69",
                                           "Population_70plus"),
                                labels = c("Age: 55 to 69",
                                           "Age: 70+"),
                                ordered = TRUE)) %>%
            st_as_sf(),
          aes(fill = Population),
          color = NA) +
  geom_sf(data = urban_types,
          aes(color = Type),
          fill = NA) +
  scale_fill_distiller(name = "Population",
                       breaks = c(20, 50, 150, 400),
                       palette = "OrRd", 
                       trans = "log",
                       direction = 1) +
  scale_color_manual(name = "",
                     values = c("Urban" = "firebrick2",
                                "Suburban" = "cornflowerblue",
                                "Rural" = "forestgreen"))  +
  facet_wrap(~ Age, nrow = 2) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 7),
        legend.position = "bottom",
        legend.text = element_text(size = 8))
```

Methods and Data
============

# Data

@@

- Urban, suburban, and rural boundary definitions from the City of Hamilton\footnote{\url{https://open.hamilton.ca/}}
- Population and median total household income for 2016 by Dissemination Area (DA) boundary using the `cancensus` [@vonBergmann2021cancensus] package
- Modal split by age for traffic analysis zones (TAZ) from the 2016 Transportation Tomorrow Survey (TTS)\footnote{\url{http://dmg.utoronto.ca/}}
- Locations of pilot pharmacies from public records
- Locations of three additional major chain pharmacies for scenario testing
- Residential units by land parcel

# Methods

Using the population aged 55 to 59 y.o. we first calculate the average number of people per dwelling and assign them proportionally to the dwellings by parcel. Second, median incomes and modal splits across walking, transit, and car are joined to the parcels. Although our data are from 2016 (the most recent available), we assume they are largely representative of current spatial and demographic trends in the city. Third, we use the `r5r` [@Pereira2021r5r] package to calculate the travel time from each parcel to all pharmacies by three modes using a cutoff value of 180 min and a maximum walking distance of 10,000 m. Once we obtained travel time tables with population, proportion of trips by mode, and income information, we calculated the expected travel time $ett$ from each parcel $i$ to a pharmacy $j$ as follows:
$$
ett_i = p^c_i \min(tt^c_{ij}) + p^t_i \min(tt^t_{ij}) + p^w_i \min(tt^w_{ij})
$$
\noindent where $p^k_i$ is the proportion of trips by mode $k$ in the TAZ of parcel $i$, and $tt^k_{ij}$ is the vector of travel times from parcel $i$ to the pharmacies. In other words, the expected travel time is the weighted sum of travel times to the nearest pharmacy, with the weights given by the expected modal split in the TAZ.

@@

For comparison purposes we consider a scenario with some modest additions to the list of pharmacies in the provincial pilot. We repeat the analysis, but include the three urban sites shown in white circles in Figure \ref{fig:pharmacies-and-regions}. The results of this scenario appear in the last two columns of Table \ref{tab:distribution-results} and the two right panels of Figure \ref{fig:results}. While all income groups benefit from the addition of these three sites with shorter mean trip durations, the most remarkable difference is the large reduction in the disparities between residents in DAs with different levels of income. The top-right panel of Figure \ref{fig:results} shows that the distribution of expected travel time is now more in line for all income groups, even if the bottom two income quintiles still have somewhat wider spreads. Unsurprisingly, the addition of three urban vaccination sites does not have a large impact for rural residents.

The results indicate that the locations chosen by the province for the pilot vaccination program do not serve well urban or rural residents of the city, and there are some important questions regarding equity of access to the program, with a disproportionate burden in the cost of travel falling on lower income urban populations and rural populations. Selection of three sensible urban locations does much to alleviate disparities in the burden of transportation. On the other hand, because there are not many candidate locations in rural parts of the city, increasing access for these residents likely will involve an expansion of the existing mobile vaccination pop-up clinic program\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.

@@

@@

```{r figure-results, echo=FALSE, warning=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:results}Distribution of expected travel time for different population groups"}
grid.arrange(ttm_income_baseline + 
               ggtitle("By income: Pilot Program") + 
               labs(x = "") + 
               xlim(c(0,40)) +
               theme(title = element_text(size = 8), 
                     axis.text = element_text(size = 7), 
                     legend.position = "none"), 
             ttm_income + 
               ggtitle("By income: Scenario") + 
               labs(x = "", y = "") + 
               xlim(c(0,40)) +
               theme(title = element_text(size = 8), 
                     axis.text = element_text(size = 7)), 
             ttm_region_baseline + 
               ggtitle("By region: Pilot Program") + 
               xlim(c(0,40)) +
               theme(title = element_text(size = 8),
                     axis.title = element_text(size = 8), 
                     axis.text = element_text(size = 7), 
                     legend.position = "none"), 
             ttm_region + 
               ggtitle("By region: Scenario") + 
               labs(y = "") +
               xlim(c(0,40)) +
               theme(title = element_text(size = 8),
                     axis.title = element_text(size = 8), 
                     axis.text = element_text(size = 7)), 
             nrow = 2)
```

>>>>

Questions
==========================

@@

Critics were swift to point out that the list of pharmacies approved for Hamilton by the province were mostly located in lower density parts of the city that are not well serviced by transit and are difficult to reach by foot\footnote{See inter alia: \url{https://twitter.com/RyanMcGreal/status/1378027149790224386?s=20} and \url{https://twitter.com/NrinderWard3/status/1378679195514060801?s=20}.}. Indeed, as seen in Figure \ref{fig:pharmacies-and-regions}, a vast majority of the pharmacies are in suburban Hamilton. The issue is less clear-cut when we consider that Hamilton's older population skews suburban (see Figure \ref{fig:population-map}). Given the target demographic for the program, it is possible that suburban sites could be convenient for mature adults and the young old: the population aged 55 to 69 in Hamilton is approximately `r prettyNum(round((table_comparison[7,2]/1000), 2) * 1000, big.mark = ",")` suburban, `r prettyNum(round((table_comparison[8,2]/1000), 2) * 1000, big.mark = ",")` urban, and only `r prettyNum(round((table_comparison[6,2]/1000), 2) * 1000, big.mark = ",")` rural. Nevertheless, the selection of sites by the province raises some important questions\footnote{The decision-making process to select these sites appears to have been opaque, and the Mayor of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}. As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may lead to inequities in access.

@@

```{r pharmacies-map, echo=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:pharmacies-and-regions}Regions with the City of Hamilton; the location of pharmacies in pilot is shown (black triangles) along with urban locations for scenario analysis (white circles)."}
#map_pharmacies <-
ggplot() +
  geom_sf(data = urban_types,
          aes(fill = Type),
          color = "lightgray") +
  geom_sf(data = pharmacy_locations,
          aes(shape = Type,
              color = Type),
          size = 2.5,
          show.legend = FALSE) +
  scale_color_manual(values = c("Candidate" = "white",
                                "Pilot" = "black")) +
  scale_fill_manual(values = c("Urban" = "firebrick2",
                               "Suburban" = "cornflowerblue",
                               "Rural" = "forestgreen")) +
  guides(fill = guide_legend(title = "")) + 
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 7),
        legend.position = "bottom",
        legend.text = element_text(size = 8))
```

```{r save-pharmacies-map-1, include=FALSE}
ggplot() +
  geom_sf(data = urban_types,
          aes(fill = Type),
          color = "lightgray") +
  geom_sf(data = pharmacy_locations %>%
            filter(Type == "Pilot"),
          aes(shape = Type,
              color = Type),
          size = 2.5,
          show.legend = FALSE) +
  scale_shape_manual(values = c("Candidate" = 16,
                                "Pilot" = 17)) +
  scale_color_manual(values = c("Candidate" = "white",
                                "Pilot" = "black")) +
  scale_fill_manual(values = c("Urban" = "firebrick2",
                               "Suburban" = "cornflowerblue",
                               "Rural" = "forestgreen")) +
  guides(fill = guide_legend(title = "")) + 
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 7),
        legend.position = "bottom",
        legend.text = element_text(size = 8))

ggsave("images/pharmacies-map-1.jpeg")
```

```{r save-pharmacies-map-2, include=FALSE}
ggplot() +
  geom_sf(data = urban_types,
          aes(fill = Type),
          color = "lightgray") +
  geom_sf(data = pharmacy_locations,
          aes(shape = Type,
              color = Type),
          size = 2.5,
          show.legend = FALSE) +
  scale_shape_manual(values = c("Candidate" = 16,
                                "Pilot" = 17)) +
  scale_color_manual(values = c("Candidate" = "white",
                                "Pilot" = "black")) +
  scale_fill_manual(values = c("Urban" = "firebrick2",
                               "Suburban" = "cornflowerblue",
                               "Rural" = "forestgreen")) +
  guides(fill = guide_legend(title = "")) + 
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 7),
        legend.position = "bottom",
        legend.text = element_text(size = 8))

ggsave("images/pharmacies-map-2.jpeg")
```

```{r population-map, echo=FALSE, warning=FALSE, fig.height=8, fig.align="center", fig.cap="\\label{fig:population-map}Distribution of population aged 55+ in the City of Hamilton."}
ggplot() +
  geom_sf(data = data_da_2016 %>%
            filter(`Region Name` == "Hamilton") %>%
            dplyr::select(Population_55to69, Population_70plus) %>%
            pivot_longer(cols = -geometry, names_to = "Age", values_to = "Population") %>%
            mutate(Age = factor(Age,
                                levels = c("Population_55to69",
                                           "Population_70plus"),
                                labels = c("Age: 55 to 69",
                                           "Age: 70+"),
                                ordered = TRUE)) %>%
            st_as_sf(),
          aes(fill = Population),
          color = NA) +
  geom_sf(data = urban_types,
          aes(color = Type),
          fill = NA) +
  scale_fill_distiller(name = "Population",
                       breaks = c(20, 50, 150, 400),
                       palette = "OrRd", 
                       trans = "log",
                       direction = 1) +
  scale_color_manual(name = "",
                     values = c("Urban" = "firebrick2",
                                "Suburban" = "cornflowerblue",
                                "Rural" = "forestgreen"))  +
  facet_wrap(~ Age, nrow = 2) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 7),
        legend.position = "bottom",
        legend.text = element_text(size = 8))
```

```{r save-population-map, include=FALSE}
ggplot() +
  geom_sf(data = data_da_2016 %>%
            filter(`Region Name` == "Hamilton") %>%
            dplyr::select(Population_55to69, Population_70plus) %>%
            pivot_longer(cols = -geometry, names_to = "Age", values_to = "Population") %>%
            mutate(Age = factor(Age,
                                levels = c("Population_55to69",
                                           "Population_70plus"),
                                labels = c("Age: 55 to 69",
                                           "Age: 70+"),
                                ordered = TRUE)) %>%
            st_as_sf(),
          aes(fill = Population),
          color = NA) +
  geom_sf(data = urban_types,
          aes(color = Type),
          fill = NA) +
  scale_fill_distiller(name = "Population",
                       breaks = c(20, 50, 150, 400),
                       palette = "OrRd", 
                       trans = "log",
                       direction = 1) +
  scale_color_manual(name = "",
                     values = c("Urban" = "firebrick2",
                                "Suburban" = "cornflowerblue",
                                "Rural" = "forestgreen"))  +
  facet_wrap(~ Age, nrow = 2) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 7),
        legend.position = "bottom",
        legend.text = element_text(size = 8))

ggsave("images/population-map.jpeg")
```

Methods
============

@@

- Urban, suburban, and rural boundary definitions from the City of Hamilton\footnote{\url{https://open.hamilton.ca/}}
- Population and median total household income for 2016 by Dissemination Area (DA) boundary using the `cancensus` package [@vonBergmann2021cancensus]
- Modal split by age for traffic analysis zones (TAZ) from the 2016 Transportation Tomorrow Survey (TTS)\footnote{\url{http://dmg.utoronto.ca/}}
- Locations of pilot pharmacies from public records
- Locations of three additional major chain pharmacies for scenario testing
- Residential units by land parcel

Using the population aged 55 to 59 y.o. we first calculate the average number of people per dwelling and assign them proportionally to the dwellings by parcel. Second, median incomes and modal splits across walking, transit, and car are joined to the parcels. Although our data are from 2016 (the most recent available), we assume they are largely representative of current spatial and demographic trends in the city. Third, we use the `r5r` [@Pereira2021r5r] package to calculate the travel time from each parcel to all pharmacies by three modes using a cutoff value of 180 min and a maximum walking distance of 10,000 m. Once we obtained travel time tables with population, proportion of trips by mode, and income information, we calculated the expected travel time $ett$ from each parcel $i$ to a pharmacy $j$ as follows:
$$
ett_i = p^c_i \min(tt^c_{ij}) + p^t_i \min(tt^t_{ij}) + p^w_i \min(tt^w_{ij})
$$
\noindent where $p^k_i$ is the proportion of trips by mode $k$ in the TAZ of parcel $i$, and $tt^k_{ij}$ is the vector of travel times from parcel $i$ to the pharmacies. The expected travel time is thus the weighted sum of travel times to the nearest pharmacy, with the weights given by the expected modal split in the TAZ.

@@

For comparison purposes we consider a scenario with some modest additions to the list of pharmacies in the provincial pilot. We repeat the analysis, but include the three urban sites shown in white circles in Figure \ref{fig:pharmacies-and-regions}. The results of this scenario appear in the last two columns of Table \ref{tab:distribution-results} and the two right panels of Figure \ref{fig:results}. While all income groups benefit from the addition of these three sites with shorter mean trip durations, the most remarkable difference is the large reduction in the disparities between residents in DAs with lower levels of income. The top-right panel of Figure \ref{fig:results} shows that the distribution of expected travel time is now more in line for all income groups, even if the bottom two income quintiles still have somewhat wider spreads. Unsurprisingly, the addition of three urban vaccination sites does not have a large impact for rural residents.

The results indicate that the locations chosen by the province for the pilot vaccination program do not serve well urban or rural residents of the city, and there are some important questions regarding equity of access to the program, with a disproportionate burden in the cost of travel falling on lower income urban populations and rural populations. Selection of three sensible urban locations does much to alleviate disparities in the burden of transportation. On the other hand, because there are not many candidate locations in rural parts of the city, increasing access for residents there likely necessitates an expansion of the existing mobile vaccination pop-up clinic program\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.

@@

```{r save-figure-maps-baseline, include=FALSE}
jpeg(filename = "images/baseline-maps.jpeg")
grid.arrange(ttm_map_baseline +
               theme(legend.title = element_text(size = 8),
                     legend.text = element_text(size = 7), 
                     axis.text = element_text(size = 7)),
             pht_map_baseline +
               theme(legend.title = element_text(size = 8),
                     legend.text = element_text(size = 7), 
                     axis.text = element_text(size = 7)),
             nrow = 2)
dev.off()

ttm_map_baseline
ggsave("images/ttm-baseline-map.jpeg")

pht_map_baseline
ggsave("images/pht-baseline-map.jpeg")
```

@@

```{r figure-results, echo=FALSE, warning=FALSE, fig.height=4, fig.align="center", fig.cap="\\label{fig:results}Distribution of expected travel time for different population groups."}
grid.arrange(ttm_income_baseline + 
               ggtitle("By income: Pilot Program") + 
               labs(x = "") + 
               xlim(c(0,40)) +
               theme(title = element_text(size = 8), 
                     axis.text = element_text(size = 7), 
                     legend.position = "none"), 
             ttm_income + 
               ggtitle("By income: Scenario") + 
               labs(x = "", y = "") + 
               xlim(c(0,40)) +
               theme(title = element_text(size = 8), 
                     axis.text = element_text(size = 7)), 
             ttm_region_baseline + 
               ggtitle("By region: Pilot Program") + 
               xlim(c(0,40)) +
               theme(title = element_text(size = 8),
                     axis.title = element_text(size = 8), 
                     axis.text = element_text(size = 7), 
                     legend.position = "none"), 
             ttm_region + 
               ggtitle("By region: Scenario") + 
               labs(y = "") +
               xlim(c(0,40)) +
               theme(title = element_text(size = 8),
                     axis.title = element_text(size = 8), 
                     axis.text = element_text(size = 7)), 
             nrow = 2)
```

```{r save-figure-results, include=FALSE}
jpeg(filename = "images/figure-results.jpeg")
grid.arrange(ttm_income_baseline + 
               ggtitle("By income: Pilot Program") + 
               labs(x = "") + 
               xlim(c(0,40)) +
               theme(title = element_text(size = 8), 
                     axis.text = element_text(size = 7), 
                     legend.position = "none"), 
             ttm_income + 
               ggtitle("By income: Scenario") + 
               labs(x = "", y = "") + 
               xlim(c(0,40)) +
               theme(title = element_text(size = 8), 
                     axis.text = element_text(size = 7)), 
             ttm_region_baseline + 
               ggtitle("By region: Pilot Program") + 
               xlim(c(0,40)) +
               theme(title = element_text(size = 8),
                     axis.title = element_text(size = 8), 
                     axis.text = element_text(size = 7), 
                     legend.position = "none"), 
             ttm_region + 
               ggtitle("By region: Scenario") + 
               labs(y = "") +
               xlim(c(0,40)) +
               theme(title = element_text(size = 8),
                     axis.title = element_text(size = 8), 
                     axis.text = element_text(size = 7)), 
             nrow = 2)
dev.off()

ttm_income_baseline + 
               ggtitle("By income: Pilot Program") + 
               labs(x = "") + 
               xlim(c(0,40))
ggsave("images/ttm-income-pilot.jpeg")

ttm_region_baseline + 
  ggtitle("By region: Pilot Program") + 
  xlim(c(0,40))
ggsave("images/ttm-region-pilot.jpeg")

ttm_income +
  ggtitle("By income: Scenario") + 
  labs(x = "", y = "") + 
  xlim(c(0,40))
ggsave("images/ttm-income.jpeg")

ttm_region + 
  ggtitle("By region: Scenario") + 
  labs(y = "") +
  xlim(c(0,40))
ggsave("images/ttm-region.jpeg")
```
